MOMS-PI: Regression on diversity

### 2022-01-10 notebook/_posts/2021-12-15-momspi-regression-diversity/_cut.Rmd
Revise MOMSPI regression analysis
- Removed rank regression and Gamma-Poisson regression (kept in
  _cut.Rmd) until have time to better understand these
- Adjusted zero-replacement for the lm regression
- General refactoring for clarity
- Improved plots

---
title: "MOMS-PI: Regression on diversity"
description: |
author:
  - name: Michael R. McLaren
    url: {}
categories:
  - ref:brooks2015thet
  - ref:fettweis2019thev
date: 2021-12-15
draft: false
bibliography: ../../../main.bib
output:
  distill::distill_article:
    self_contained: false
    dev: svg
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  cache = TRUE,
  autodep = TRUE,
  cache.comments = FALSE,
  dpi = 300
)
```

# Setup

## R setup

```{r libraries}
# Tools for microbiome data
library(speedyseq)
# Tools for general purpose data manipulation and plotting
library(tidyverse)
library(fs)
# ggplot helpers
library(ggbeeswarm)
library(cowplot)
library(patchwork)
library(scales)
# stats helpers
# library(broom)

library(metacal); packageVersion("metacal")
```

```{r ggplot}
theme_set(theme_cowplot())

colors_brooks <- c(
  "Atopobium_vaginae" = "#009E73",
  "Gardnerella_vaginalis" = "#56B4E9",
  "Lactobacillus_crispatus" = "#D55E00",
  "Lactobacillus_iners" = "#505050",
  "Prevotella_bivia" = "#0072B2",
  "Sneathia_amnii" = "#CC79A7",
  "Streptococcus_agalactiae" = "#E69F00")

scale_y_custom <- scale_y_continuous(
    trans = 'log10',
    breaks = trans_breaks('log10', function(x) 10^x),
    labels = trans_format('log10', math_format(10^.x))
  )
```

# SETUP TODO

# Regression analysis

NOTE: Consider saving output so can be used in other notebook files

## Classify samples into high and low diversity groups

TODO: Consider possible better way to do calibration

Compute Shannon and Inverse-Simpson diversity indices and add to the sample metadata.
`Phyloseq::estimate_richness()` will complain about the lack of singletons, but we can ignore this since we aren't using species richness estimators.
I will exponentiate the Shannon index to convert it to the q1 Hill number so that (like Inverse Simpson) it has the scale of number of species.

```{r}
div <- list(
  'obs' = momspi_obs,
  'cal' = momspi_cal
) %>%
  map(estimate_richness, measures = c('Shannon', 'InvSimpson')) %>%
  map_dfr(as_tibble, rownames = '.sample', .id = 'type') %>%
  transmute(
    .sample, type,
    div_q1 = exp(Shannon),
    div_q2 = InvSimpson,
  ) %>%
  pivot_longer(starts_with('div_'), names_to = 'index', values_to = 'diversity')
div_wide <- div %>%
  pivot_wider(values_from = diversity, names_from = c('index', 'type')) 
```

```{r}
div_wide %>%
  mutate(across(starts_with('div'), log2)) %>%
  ggplot() +
  geom_point(aes(x = .panel_x, y = .panel_y), alpha = 0.5) +
  ggforce::geom_autodensity() +
  geom_bin2d(aes(x = .panel_x, y = .panel_y)) +
  ggforce::facet_matrix(vars(starts_with('div')), layer.diag = 2, layer.lower = 3)
```

The observed (Log) order-1 and order-2 diversity show a tight correlation with each other, as do the calibrated diversities, while the correlation is much lower when comparing observed and calibrated numbers.
The reason is likely that calibration will tend to increase the diversity of samples that are observed to be dominated by a high-efficiency taxon (like L. iners) and decrease the diversity of samples dominated by a low efficiency tason (like G. vaginalis).

Shannon diversity of the observed profiles is what is most often considered in vaginal microbiome studies, so let's use this metric to partition the samples.

```{r}
div_wide %>%
  ggplot(aes(x = div_q1_obs)) +
  scale_x_log10() +
  annotate('rect', xmin = 0, xmax = 1.5, ymin = -Inf, ymax = Inf, 
    fill = 'blue', alpha = 0.3) +
  annotate('rect', xmin = 1.5, xmax = 3, ymin = -Inf, ymax = Inf, 
    fill = 'grey', alpha = 0.3) +
  annotate('rect', xmin = 3, xmax = Inf, ymin = -Inf, ymax = Inf, 
    fill = 'green', alpha = 0.3) +
  geom_histogram() +
  geom_vline(xintercept = c(1.5, 3), color = 'black')
```

I'll pick the first mode as the 'Low diversity' group, treat the second mode as 'Medium', and the tail as 'High'.

```{r}
div_wide <- div_wide %>%
  mutate(
    div_group = case_when(
      div_q1_obs < 1.5 ~ 'Low',
      div_q1_obs > 3 ~ 'High',
      TRUE ~ 'Medium'
    ) %>% factor(levels = c('Low', 'Medium', 'High'))
  )
```

## Understanding the diversity-efficiency relationship

```{r}
mean_eff3 <- mean_eff2 %>%
  left_join(
    div,
    # div %>% pivot_wider(names_from = type, values_from = c(Shannon, InvSimpson)),
    by = '.sample'
  )
```

```{r}
mean_eff3 %>%
  ggplot(aes(diversity, mean_efficiency, color = top_plot)) +
  facet_grid(type~index) +
  scale_x_log10() +
  scale_y_log10() +
  geom_point(alpha = 0.6) +
  stat_smooth(aes(group = 'none'), color = 'black') +
  scale_color_manual(
    values = colors_top,
    labels = function(x) str_replace_all(x, '_', ' ')
  )
```

```{r}
sam_div <- momspi_div %>%
  sample_data %>%
  as_tibble
```

```{r}
sam_div %>%
  ggplot(aes(y = div_group, x = mean_efficiency)) +
  scale_x_log10() +
  ggdist::stat_slab()
```

We can see that the mean efficiency is 3-10X lower in the High diversity relative to the low diversity group.

```{r}
lm(log(mean_efficiency) ~ div_group, data = sam_div) %>% summary
```

## Regression setup

To simplify interpretation of some of the regression models, let's rarefy samples to 10K reads (we'll be dropping some samples).

```{r}
momspi_rare <- momspi %>% 
  rarefy_even_depth(
    sample.size = 1e4, 
    rngseed = 42,
    replace = TRUE, # for speed, might want to change
    trimOTUs = FALSE
  )
```

Add the diversity grouping information and the offset terms for the Gamma-Poisson (Negative Binomial) regression.

```{r}
momspi_div <- momspi_rare %>%
  left_join_sample_data(mean_eff, by = '.sample') %>%
  left_join_sample_data(div_wide, by = '.sample') %>%
  mutate_sample_data(
    gp_offset_obs = log(sample_sum),
    gp_offset_cal = log(sample_sum) - log(mean_efficiency),
  )
```

We'll use the criterion from Fettweis et al to pick the taxa to test.
Note, we may not get the same exact taxa as them since we're using a larger set of women.
Note, the sample sums used in the offset were calculated for the full species set, so we are still analzying log proportions with respect to the full species set.

```{r}
taxa_to_test <- momspi %>%
  transform_sample_counts(close_elts) %>%
  filter_taxa2(~ mean(. >= 1e-2) >= 0.05 | mean(. >= 1e-3) >= 0.15) %>%
  taxa_names %>%
  print
```

## Linear regression on log proportion

Do this first to confirm as expected from theory.

```{r}
stat_test <- function(ps) {
  ps %>%
    transform_sample_counts(close_elts) %>%
    filter_sample_data(div_group %in% c('Low', 'High')) %>%
    prune_taxa(taxa_to_test, .) %>%
    as_tibble %>%
    with_groups(.otu, nest) %>%
    mutate(
      fit = map(data,
        ~lm(log(.abundance) ~ div_group, data = .x)
      ),
    ) %>%
    select(-data)
  # %>% unnest(fit)
}
res_div_lm <- list(
  'Observed' = momspi_div %>% transform_sample_counts(~. + 0.3),
  'Calibrated' = momspi_div %>% transform_sample_counts(~. + 0.3) %>% calibrate(bias_all_vec)
) %>%
  map_dfr(stat_test, .id = 'type')
res_div_lm_tidy <- res_div_lm %>%
  mutate(across(fit, map, broom.mixed::tidy)) %>%
  unnest(fit) %>% 
  filter(term != '(Intercept)')
```

```{r}
# x should be the outout of a tidier
plot_estimates <- function(x) {
  err <- x %>%
    select(name = type, .otu, value = estimate) %>%
    pivot_wider() %>%
    mutate(
      # error = Observed - Calibrated, # will be same for everything
      # error_abs = abs(error),
      # # error_rel = abs(error / Calibrated), # only makes sense if not a sign error
      error_sign = sign(Observed) == sign(Calibrated),
      sign = case_when(
        sign(Observed) > 0 & sign(Calibrated) > 0 ~ 'Positive',
        sign(Observed) < 0 & sign(Calibrated) < 0 ~ 'Negative',
        TRUE ~ 'Depends',
      )
    ) %>%
    select(.otu, error_sign, sign)

  p1 <- x %>%
    left_join(err, by = '.otu') %>%
    ggplot(aes(y = type, x = estimate, fill = sign)) +
    scale_fill_brewer(type = 'qual', palette = 2) +
    geom_vline(xintercept = 0, color = 'grey') +
    ggdist::geom_dots()
  p2 <- x %>%
    mutate(
      .otu_fct = fct_reorder(.otu, estimate, .fun = min)
    ) %>%
    ggplot(aes(y = .otu_fct, x = estimate, color = type)) +
    scale_color_brewer(type = 'qual', palette = 7) +
    geom_vline(xintercept = 0, color = 'grey') +
    geom_pointrange(
      aes(
        xmin = estimate - 2*std.error,
        xmax = estimate + 2*std.error
      ))
  p1 / p2 +
    plot_layout(ncol = 1, heights = c(0.2, 1))
}
plot_estimates(res_div_lm_tidy)
```

Consider doing bootstrap CIs.

```{r, eval = F}
lm_boot <- function(.data, ...) {
  fit <- lm(log(.abundance) ~ div_group, data = .data)
  car::Boot(fit, ...)
}
x <- momspi_div %>% transform_sample_counts(~. + 0.3) %>% 
  filter_sample_data(div_group %in% c('Low', 'High')) %>%
  filter_tax_table(.otu == 'Gardnerella_vaginalis') %>% 
  as_tibble
fit <- lm(log(.abundance) ~ div_group, data = x)
bt <- car::Boot(fit)
bt %>% broom::tidy()
broom::tidy(bt, conf.int = TRUE)
confint(bt, type = 'perc') %>% broom::tidy()
```

@@

Similar to Wilcoxin test when there are no ties; not sure about with ties.

```{r}
stat_test <- function(ps) {
  ps %>%
    transform_sample_counts(close_elts) %>%
    filter_sample_data(div_group %in% c('Low', 'High')) %>%
    prune_taxa(taxa_to_test, .) %>%
    as_tibble %>%
    with_groups(.otu, nest) %>%
    mutate(
      fit = map(data,
        ~lm(
          rank(.abundance, ties.method = 'average')/n() ~ div_group, 
          data = .x
        )
      ),
    ) %>%
    select(-data)
  # %>% unnest(fit)
}
res_div_rank <- list(
  'Observed' = momspi_div %>% transform_sample_counts(~. + 0.3),
  'Calibrated' = momspi_div %>% transform_sample_counts(~. + 0.3) %>% calibrate(bias_all_vec)
) %>%
  map_dfr(stat_test, .id = 'type')
res_div_rank_tidy <- res_div_rank %>%
  mutate(across(fit, map, broom.mixed::tidy)) %>%
  unnest(fit) %>% 
  filter(term != '(Intercept)')
```

@@

@@

>>>>

# Regression - new

Let's create a data frame with the abundances and offsets for the different regressions.
For the analyses that do not model the counting process, we'll adjust the proportions to the Dirichlet posterior mean, with the prior set to the average proportions across all samples.

```{r}
taxa_mean_prop <- momspi_div %>%
  otu_table %>%
  transform_sample_counts(close_elts) %>%
  orient_taxa(as = 'cols') %>%
  as('matrix') %>%
  colMeans
stopifnot(sum(taxa_mean_prop) == 1)
prior_vec <- taxa_mean_prop * ntaxa(momspi)
stopifnot(identical(length(prior_vec), ntaxa(momspi)))
# Note the need for the seq_along trick to get phyloseq to allow this
# adjustment
momspi_div_adj <- momspi_div %>% 
  transform_sample_counts(~ prior_vec[seq_along(.x)] + .x) %>%
  transform_sample_counts(close_elts) %>%
  otu_table %>%
  as_tibble %>%
  rename(prop_adj = .abundance)
```

```{r}
x <- momspi_div %>%
  as_tibble %>%
  rename(count = .abundance) %>%
  with_groups(.sample, mutate,
    prop_raw = close_elts(count),
  ) %>%
  filter(.otu %in% taxa_to_test) %>%
  left_join(by = c('.sample', '.otu'), momspi_div_adj) %>%
  left_join(bias_all %>% select(.otu, efficiency), by = '.otu') %>%
  mutate(
    # Offset for log link when bias is ignored
    offset_obs = log(sample_sum),
    # Offset for log link accounting for bias
    offset_cal = log(sample_sum) + log(efficiency / mean_efficiency),
  ) %>%
  # Calibrated proportions
  with_groups(.sample, mutate,
    across(starts_with('prop_'), .names = '{.col}_cal', 
      ~close_elts(. / efficiency))
  )
all.equal(x$prop_raw_cal * x$sample_sum / exp(x$offset_cal), , x$prop_raw)

stopifnot()
```

@@

Similar to Wilcoxin test; approximately equivalent when there are no ties.

```{r}
stat_test <- function(ps) {
  ps %>%
    transform_sample_counts(close_elts) %>%
    filter_sample_data(div_group %in% c('Low', 'High')) %>%
    prune_taxa(taxa_to_test, .) %>%
    as_tibble %>%
    with_groups(.otu, nest) %>%
    mutate(
      fit = map(data,
        ~lm(
          rank(.abundance, ties.method = 'average')/n() ~ div_group, 
          data = .x
        )
      ),
    ) %>%
    select(-data)
  # %>% unnest(fit)
}
res_div_rank <- list(
  'Observed' = momspi_div_prop_adj,
  'Calibrated' = momspi_div_prop_adj %>% calibrate(bias_all_vec)
) %>%
  map_dfr(stat_test, .id = 'type')
res_div_rank_tidy <- res_div_rank %>%
  mutate(across(fit, map, broom.mixed::tidy)) %>%
  unnest(fit) %>% 
  filter(term != '(Intercept)')
```

@@

NOTE: I think the results where quite different when zeros are replaced with the same value for each taxon (????)

Try the same thing without zero replacement,

```{r}
stat_test <- function(ps) {
  ps %>%
    transform_sample_counts(close_elts) %>%
    filter_sample_data(div_group %in% c('Low', 'High')) %>%
    prune_taxa(taxa_to_test, .) %>%
    as_tibble %>%
    with_groups(.otu, nest) %>%
    mutate(
      fit = map(data,
        ~lm(
          rank(.abundance, ties.method = 'average')/n() ~ div_group, 
          data = .x
        )
      ),
    ) %>%
    select(-data)
  # %>% unnest(fit)
}
res_div_rank <- list(
  'Observed' = momspi_div,
  'Calibrated' = momspi_div %>% calibrate(bias_all_vec)
) %>%
  map_dfr(stat_test, .id = 'type')
res_div_rank_tidy <- res_div_rank %>%
  mutate(across(fit, map, broom.mixed::tidy)) %>%
  unnest(fit) %>% 
  filter(term != '(Intercept)')
```

This is very mysterious. Needs more investigation before we report these results.

@@

HERE. Don't have time to mess with this for now. Set eval to false for now.
